Bayesian exponential random graph modeling of KWL Burial network

### 2020-11-20 analysis/paper/paper.Rmd
editing methods and results

<!-- our hypothesis --> 
Our pericolonial impact hypothesis is that rare, high quality grave goods accumulated in only a few individuals' burials after the European presence in the 17th century, because of increased social inequality at Kiwulan due to differential access to these goods. We can identify these high value trade goods had been from historical documents written by the Europeans and the Han Chinese [@Borao2009; @LiandWu2006]. A related effect of the European presence in northeastern Taiwan may have been manipulation of the European colonial image by ambitious local Indigenous individuals for building their personal status and power. If social inequality gradually increased in Kiwulan as we hypothesize, then we expect to observe a network with higher degree (popularity), less transitivity (cohesion), and strong inequality based on age difference (achieved status). To test this prediction, we use ERGM in a Bayesian framework to examine the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distribution for the priors is typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of edge to low density for both network models. For covariates based on burial attributes, such as age, gender, and wealth index, we specified a vague prior information that follows a normal distribution with mean at 0 and standard deviation at 5 (i.e. N(0, 5)) for both models. For physical distance between burials, we set a negative covariate effect (N(-1, 5)) to infer the kinship-based proximity. To evaluate our anthropological assumption about increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for social inequality, especially for gwesp and gwedegree representing transitivity and popularity. We set the priors to higher transitivity (N(5, 7)), lower popularity (N(4, 7)), and higher covariate effect based on the ritual (N(2, 5)) for network before European contact to indicate less social inequality. On the contrary, we set the prior for the network after the European arrival to lower transitivity (N(4, 8)), and higher popularity (N(8, 8)) to illustrate an increased social inequality. This prior information derives from the theory about horizontal hierarchy, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate mode at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set differently with N(-17, 7) for earlier network and N(-20, 6) for later network. It should be noted the idea of those settings is to reflect the relative differences between two networks, and the standard deviation was adjusted in terms of different networks to indicate uncertainty.

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
To assess MCMC analysis, we first evaluated three diagnostics visual summaries, including density plots, trace plots, and autocorrelation plots [@Hamra2013]. Those plots provide an informal way to diagnose model convergence by identifying any unexpected peaks or shapes in distribution. Then we access output from our two  models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. Our Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. We set of 10000 network graphs simulated from the estimated posterior distribution in ERGMs. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network.

@@

The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at xxx to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

<!--posterior stats for pre-E-->
We compared estimated statistics from the posterior distributions, networks before and after the European presence, to examine their differences in network structure (Table XX). For the former network, the nodal covariates, age-homophily and ritual-homophily, have significant effects on the relations between burials. The age homophiligy shows a negative effect that indicates burials of the same age tend to not share similar prestige goods, while the ritual-homophily demonstrates the positive effect that means burials with ritual pottery tend to form relations. Despite positive value from some covariates, such as sex-homophily and burial value-homophily, they do not show a significant tendency to forming relations duo to value of zero in a confidence interval. Similarly, the dyadic covariate, physical distance, shows positive effects but not significantly. For the endogenous network effects, both gwesp and gwdegree present significant positive effects. The positive value of gwesp suggests a tendency of burials to be clustered in closed transitive structures, while positive gwdegree indicates a tendency toward centralization that means the presence of a limited number of burials with more relations.

<!--posterior stats for for post-E and their comparison-->
For the network after the arrival of Europeans, there are no significant effects of nodal and dyadic covariates, but we still have a clue of possible tendency. The nodal covariates of age-homophily, sex-homophily, ritual-homophily, and dyadic covariate of physical distance all show positive effects, while the burial value-homophily has a negative effect. We also found that the burials in the same wealth level may tend not to form relations. Similar to the earlier network, the endogenous network variables, gwesp and gwdegree, all demonstrate positive effects significantly. However, there are differences in values of positive effects between two networks. The later network has a smaller gwesp effect and a larger gwedegree effect that may suggest a less tendency toward clustering but more tendency toward centralization after the European presence. Both posterior estimates presents symmetric distributions because of the posterior means are close to posterior medians (Table XX).

@@

<!--comparison between two networks using bootstrap--> 
To understand whether the differences in posterior distributions are affected by different network sizes, we conducted a sensitivity test using vertex bootstrap inference on networks that allows a comparison for multiple networks of different sizes [@Chen2019]. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) of a network to quantify standard errors in the network statistics estimation of interest. This enables the evaluation of uncertainty for networks and the comparison between networks by examining their confidence interval for the network population. We set the resampling size to 1,000 for network statistics, including density, clustering, and degree, for two networks. The result shows that the 95% confidence interval of each network statistic contains the observed network of both periods, which indicates our network is less sensitive to sample size. This suggests a limited sample size effect on both models. In addition, since the prior setting will have impacts on the posterior, we compared the posterior estimates with the same prior for all parameters, and those with the informative prior based on our assumptions for some parameters. In general, there are no significant changes in the posterior estimates using different sets of priors. But we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation.

>>>>

<!-- our hypothesis --> 
Our pericolonial impact hypothesis is that rare, high quality grave goods accumulated in only a few individuals' burials after the European presence in the 17th century, because of increased social inequality at Kiwulan due to differential access to these goods. We can identify these high value trade goods from historical documents written by Europeans and the Han Chinese [@Borao2009; @LiandWu2006]. A related effect of the European presence in northeastern Taiwan may have been manipulation of the European colonial image by ambitious local Indigenous individuals for building their personal status and power. If social inequality gradually increased in Kiwulan as we hypothesize, then we expect to observe a network with higher degree (popularity), less transitivity (cohesion), and strong inequality based on age difference (achieved status). To test this prediction, we use ERGM in a Bayesian framework to examine the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distributions for the priors are typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of the edge density parameter to low for both network models. For covariates based on burial attributes, such as age, gender, and our wealth index, we specified a vague prior that follows a normal distribution with mean at 0 and standard deviation at 5 (i.e. N(0, 5)) for both models. For physical distance between burials, we set a negative covariate effect (N(-1, 5)) to infer kinship-based proximity, e.g. stronger correlations for shorter distances. To evaluate our anthropological assumption about increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for social inequality, especially for transitivity (gwesp) and popularity (gwedegree). We set the priors to higher transitivity (N(5, 7)), lower popularity (N(4, 7)), and higher covariate effect based on ritual activity (N(2, 5)) for network before European contact to indicate less social inequality. On the contrary, we set the priors for the network after the European arrival to lower transitivity (N(4, 8)), and higher popularity (N(8, 8)) to  model an increased social inequality. This prior information derives from theory about horizontal hierarchies, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate mode at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set differently with N(-17, 7) for pre-European network and N(-20, 6) for post-European network.

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
As an informal way to diagnose model convergence, we first evaluated three diagnostic visual summaries of our MCMC output, including density plots, trace plots, and autocorrelation plots [@Hamra2013], which we include in our Supplementary Online Materials. Then we summarise output from our two models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. We set 10,000 network graphs simulated from the estimated posterior distribution in ERGMs. Our Bayesian GOF diagnostics summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions.This provides a statistical approach to check how well the estimated posterior parameter distribution, based on our hypotheses, can reproduce networks with similar general structural features of the observed networks. We then compared the distribution of our observed networks, the networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network, such as the relative importance of age and sex.

@@

The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://doi.org/xxx/xxx to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

<!--posterior stats for pre-E-->
We compared estimated statistics from the posterior distributions to examine their differences in  structure of the simulated networks (Table XX). For the pre-European network simulations, two nodal covariates, age-homophily and ritual-homophily, have significant effects on the relations between burials. The age homophily shows a negative effect that indicates burials of the same age tend to not share similar prestige goods, while the ritual-homophily demonstrates the positive effect that means burials with ritual pottery tend to form relations. Despite positive values for some covariates, such as sex-homophily and burial value-homophily, they do not show a significant tendency to form relations duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows positive effects but not significantly, indicating that physical proximity does not affect relations. For the endogenous network effects, both transitivity (gwesp) and popularity (gwdegree) present significant positive effects. The positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, while positive popularity indicates a tendency toward centralization that means the presence of a limited number of burials with many more relations than others. <!-- what is the social anthropological implication for these values -->

<!--posterior stats for for post-E and their comparison-->
For the network after the arrival of Europeans the nodal covariates of age-homophily, sex-homophily, ritual-homophily, and the dyadic covariate of physical distance all show positive effects, while the burial value-homophily has a negative effect. However, there are no significant effects for these nodal and dyadic covariates. We also found that the burials in the same wealth level tend not to form relations. Similar to the pre-European network, the endogenous network variables, transitivity (gwesp) and popularity (gwdegree), all demonstrate significant positive effects. The post-European network has a smaller transitivity effect and a larger popularity effect than the pre-European network. This may suggest a reduced tendency toward clustering but higher tendency toward centralization after the European presence. Both posterior estimates present symmetric distributions, with the posterior means located close to posterior medians (Table XX).

@@

<!-- comparison between two networks using patchwork bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with x burials compared to y burials. To understand if this network size different affects the posterior distributions, we conducted a sensitivity test using vertex bootstrap inference on the networks. This allows a comparison for multiple networks of different sizes [@Chen2019]. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) of a network to quantify standard errors in the network statistics of interest. This enables the evaluation of uncertainty for networks and the comparison between networks by examining their confidence interval for the network population. We set the resampling size to 1,000 networks to compute network statistics, including density, clustering, and degree, for two networks. The result shows that the 95% confidence interval of each network statistic contains the observed network of both periods, which indicates our networks are relatively robust to differences in sample size. This suggests a very small sample size effect on both models. In addition, since the prior setting will have impacts on the posterior, we compared the posterior estimates with the same prior for all parameters, and those with the informative prior based on our assumptions for some parameters. In general, there are no significant changes in the posterior estimates using different sets of priors. But we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation. <!-- re-write wither to justify why vertex bootstrap is the right method to answer this question, or use patchwork bootstrap and justify that-->
